---
title: "A Data Cleaning Example"
output: html_document
---

## Objective: 

Clean a dataset with single column containing multiple variables as delimited `key-value` pairs.

## Solution:

This can be done using `tidyr` functions `spread`, `gather`, `separate`, etc.

## Code and Explanation:

Creating a test dataset with multiple variables of interest -

```{r clean}
var1_text = c("Sachin", "Sourav", "Rahul", "Laxman")
var2_text = c("Virat", "Jinx", "Pujara", "Rohit")
sep1 = ":"
sep2 = "|"
no_rows = 100
set.seed(9653)
d1 = data.frame(id = 1:no_rows,
            retired = sample(x = var1_text, size = 10, replace = TRUE),
            current = sample(x = var2_text, size = 10, replace = TRUE),
            garbage = paste0("my_var", 1:no_rows),
            stringsAsFactors = FALSE)

```

Combining variable names (`keys`) with variable `values` to make it slightly unclean.

```{r unclean1}
d2 = d1
var_names = names(d1)[-1]
d2$var1_pair = paste(var_names[1], d2$retired, sep = sep1)
d2$var2_pair = paste(var_names[2], d2$current, sep = sep1)
d2$var3_pair = paste(var_names[3], d2$garbage, sep = sep1)

d2 = d2[, c("id", "var1_pair", "var2_pair", "var3_pair")]
```

Combining the `key-value` pairs for these variables into a single variable separated by a delimiter.

```{r unclean2}
d3 = d2
d3$text = NA
d3$text[4 * (1:25) - 3] = paste(d3$var1_pair[4 * (1:25) - 3], 
                                d3$var2_pair[4 * (1:25) - 3], 
                                d3$var3_pair[4 * (1:25) - 3], 
                                sep = sep2)
d3$text[4 * (1:25) - 2] = paste(d3$var2_pair[4 * (1:25) - 2], 
                                d3$var3_pair[4 * (1:25) - 2], 
                                sep = sep2)
d3$text[4 * (1:25) - 1] = paste(d3$var3_pair[4 * (1:25) - 1], 
                                d3$var2_pair[4 * (1:25) - 1], 
                                d3$var1_pair[4 * (1:25) - 1], 
                                sep = sep2)
d3$text[4 * (1:25)] = d3$var2_pair[4 * (1:25)]

d3 = d3[ , c("id", "text")]

```  

The test dataset now looks like -

```{r op1, echo = FALSE}
knitr::kable(head(d3))
```

Couple of assumptions about this dataset make this problem a non-trivial one -

1. Not all the instances (rows) contain **exactly 3** variables (`key-value` pairs).
2. The order of the variables is **not the same** i.e. `var1` not always followed by `var2` and `var2` is not always followed by `var3`. 

```{r libs, message = FALSE}
library(plyr)
library(dplyr)
library(tidyr)
library(stringr)
```

A 2 step solution: 

* step 1: Separating the `key: value pairs` into different columns i.e.
`key1: value1 | key2: value2`  ----> `key1: value1`            and      `key2: value2`

First we need to know at the most how many key value pairs could be there in a single `text` instance. Once we know the max number of vars, we can use `tidyr::separate` function to separate these pairs -

```{r cleansing1}
# Finding max number of key: val pairs in a row
len = max(str_count(string = d3$text, pattern = paste0("[",sep2,"]")))
vec_names = paste0("X", 1:(len + 1))

d2_rev = d3 %>%
  separate(col = "text", into = vec_names, sep = paste0("[",sep2,"]"), extra = "drop")
```

Result after this step -

```{r op2, echo = FALSE}
knitr::kable(head(d2_rev))
```


* step 2: Separating `value`s from `key`s i.e.
`key1: value1`  -----> `key1` and `value1`

The next step is to separate `key: val` pairs such that the `val` values are stored in corresponding `key` columns. For this, we'll first need to go from the wide version that we have now to long version of the dataset using `tidyr::gather`. After that, it's again applying `tidyr::separate` to separate `key`s from `value`s. The last step is to convert this long dataset again to a wide one using `tidyr::spread`.


```{r cleaning2}
d3_rev = d2_rev %>%
  gather(key = "temp_var", value = "kv_pair", -id, na.rm = TRUE) %>%
  select(-temp_var) %>%
  separate(col = "kv_pair", into = c("key", "val"), sep = paste0("[",sep1,"]"), extra = "drop") %>%
  spread(key = "key", value = "val")

```

Final result -

```{r op4, echo = FALSE}
knitr::kable(head(d3_rev))
```

This is very similar to the original clean dataset `d1` except for the column order and some `NA` values (generated cause of missing values we introduced while creating the test dataset, in order to make this example more generic).

```{r op_clean, echo = FALSE}
knitr::kable(head(d1))
```


