---
title: "Playing Around With Survival Analysis - 1"
output: 
  html_document: 
  fig_width: 6
  fig_height: 4

---



## Objective: 

1. Estimate survival function estimate for given time-to-event data. Here, event considered is a binary event.
2. Compare survival curve estimates for 2 segments of data.


## Solution:

We can use `survival` package to accomplish this. 

1. `Surv` function to get the data in the right format.
2. which will then be passed tp `survfit` function to compute simple Kaplan-Meier estimator.
3. `survdiff` function to test if difference between 2 survival functions is significant.


## Code:

```{r}
# install.packages("OIsurv")
# install.packages("survsim")
# install.packages("broom")
library(OIsurv)
# library(help=KMsurv)
library(dplyr)
library(broom)
library(ggplot2)
library(survsim)
```


```{r real_dataset}
# Real dataset (From KMSurv package)
# help(tongue)
data(tongue)
head(tongue)
summary(tongue)

# Analyzing just one type of tumor (Aneuploid Tumor)
tongue2 = tongue %>% 
  filter(type == 1) 

# Converting into a Surv object 
tongue2_surv <- Surv(tongue2$time, tongue2$delta)

# Getting KM estimator
tongue2_survfit = survfit(tongue2_surv ~ 1)
plot(tongue2_survfit)
glance(tongue2_survfit)

# Graphically Comparing KM estimator for 2 tumors
tongue_survfit = survfit(Surv(time = time, event = delta) ~ type, data = tongue)
plot(tongue_survfit, lty = 2:3)
legend(100, .8, c("Aneuploid", "Diploid"), lty = 2:3) 

# ggplot2 version of the plot
tongue_tidy = tidy(tongue_survfit)
mx = max(tongue_tidy$n.censor)
ggplot(tongue_tidy, aes(time, estimate, fill = strata)) + 
  geom_line() +
  geom_point(aes(shape = as.factor(n.censor)), size = 3) + 
  scale_shape_manual(values=c(NA, 1:mx))+
  geom_ribbon(aes(ymin=conf.low, ymax=conf.high), alpha=.25)

# Statistical tests for comparison
survdiff(Surv(time = time, event = delta) ~ type, data = tongue, rho = 0) # log-rank, default

# rho > 0 implies higher weight is given to initial part of the survival curve
# rho < 0 implies higher weight is given to latter part of the survival curve

```


### Simulating a dataset using `simple.surv.sim` function from `survsim` package.

* **Time-to-event:** Assumed to be following weibull distribution with p = 1.5 i.e. the mode of the time to event is closer to median then for exponential distribution (p = 1)
* **Time-to-censoring:** Assumed to be following weibull distribution with p = 10 i.e. censoring is happening mostly in the latter part of the survival curve.
* **Covariate:** 1 Binary covariate with p = .7 and beta of 0.6

```{r simulated_dataset}
set.seed(2365)
d1 = simple.surv.sim(n = 500, foltime = 1000, dist.ev = 'weibull', anc.ev = 1.5, beta0.ev = 2, # event dist - weibull with p = 1.5
                     dist.cens = 'weibull', anc.cens = 10, beta0.cens = 2.01, # censoring dist - weibull with p = 10
#                      z = list(c("weibull", 1)), #assuming independent observations i.e. no within subject correlation 
                     beta = list(c(.6)), x = list(c("bern", .7)) # 1 binary covariate
                     )

# View(d1)

summary(d1)
summary(as.data.frame(d1))

d1_survfit = survfit(Surv(time = start, time2 = stop, event = status) ~ x, data = d1)
d1_survfit = survfit(Surv(time = stop, event = status) ~ x, data = d1) # same as above

plot(d1_survfit, lty = 2:3)
legend(100, .8, c("x=0", "x=1"), lty = 2:3) 

# ggplot2 version of the plot
d1_tidy = tidy(d1_survfit)
mx = max(d1_tidy$n.censor)
ggplot(d1_tidy, aes(time, estimate, fill = strata)) + 
  geom_line() +
  geom_point(aes(shape = as.factor(n.censor)), size = 2) + 
  scale_shape_manual(values=c(NA, 1:mx))+
  geom_ribbon(aes(ymin=conf.low, ymax=conf.high), alpha=.25)

# Statistical tests for comparison
survdiff(Surv(time = stop, event = status) ~ x, data = d1, rho = 0) 

```

## References:

* http://anson.ucdavis.edu/~hiwang/teaching/10fall/R_tutorial%201.pdf
* http://www.utstat.utoronto.ca/reid/sta442f/2009/fox-survival.pdf
* http://www.jstatsoft.org/v59/i02/paper  
  
