---
title: "Playing Around With Survival Analysis"
output: html_document
  fig_width: 6
  fig_height: 4

---



Objective: 
1. Estimate survival function estimate for given time-to-event data. Here, event considered is a binary event.
2. Compare survival curve estimates for 2 segments of data.

Solution:
We can use `survival` package to accomplish this. 
1. `Surv` function to get the data in the right format 
2. which will then be passed tp `survfit` function to compute simple Kaplan-Meier estimator.
3. `survdiff` function to test if difference between 2 survival functions is significant.

Code:
```{r}
# install.packages("OIsurv")
# install.packages("survsim")
# install.packages("broom")
library(OIsurv)
library(help=KMsurv)
library(dplyr)
library(broom)
library(ggplot2)
library(survsim)
```


```{r real_dataset}
# Real dataset (From KMSurv package)
help(tongue)
data(tongue)
head(tongue)
summary(tongue)

# Analyzing just one type of tumor (Aneuploid Tumor)
tongue2 = tongue %>% 
  filter(type == 1) 

# Converting into a Surv object 
tongue2_surv <- Surv(tongue2$time, tongue2$delta)
tongue2_surv

# Getting KM estimator
tongue2_survfit = survfit(tongue2_surv ~ 1)
plot(tongue2_survfit)
glance(tongue2_survfit)

# Graphically Comparing KM estimator for 2 tumors
tongue_survfit = survfit(Surv(time = time, event = delta) ~ type, data = tongue)
glance(tongue_survfit)
plot(tongue_survfit, lty = 2:3)
legend(100, .8, c("Aneuploid", "Diploid"), lty = 2:3) 

ggplot(tidy(tongue_survfit), aes(time, estimate, fill = strata)) + 
  geom_line() +
  geom_ribbon(aes(ymin=conf.low, ymax=conf.high), alpha=.25)

# Statistical tests for comparison
survdiff(Surv(time = time, event = delta) ~ type, data = tongue, rho = 0) # log-rank, default
survdiff(Surv(time = time, event = delta) ~ type, data = tongue, rho = 1) # Peto & Peto modification of the Gehan-Wilcoxon test
survdiff(Surv(time = time, event = delta) ~ type, data = tongue, rho = -1)

# rho > 0 implies higher weight is given to initial part of the survival curve
# rho < 0 implies higher weight is given to latter part of the survival curve

```

```{r simulated_dataset}

```
Ref:

http://anson.ucdavis.edu/~hiwang/teaching/10fall/R_tutorial%201.pdf
http://www.utstat.utoronto.ca/reid/sta442f/2009/fox-survival.pdf

  
